The evolution of gene expression in the term placenta of viviparous mammals

### 2016-01-29 manuscript/manuscript.Rnw
make pictures slightly smaller
fix plot colors and a few more fixmes

\newcommand{\genetreeexplanation}[1]{Gene tree and heatmap of
  expression of genes related to #1. The gene tree was obtained from
  Ensembl~\cite{Cunningham.ea2015:Ensembl2015} and genes from species
  not included in this study or having expression by \ac{FPKM} less
  than 1 were excluded from the plotted tree. Heatmap cells in grey
  are from species where there was no expression of the relevant
  transcript (or a species where that transcript does not apply).}

@@

<<plot.defaults,echo=FALSE,results="hide">>=
grad.def <- list(low = "#132B43", high = "#56B1F7", space = "Lab", na.value = "gray50")

@@

Fetal tissue was collected from the placenta of eight mammalian
species (\human [human], \bonobo [bonobo], \dog [domestic dog], \cow
[domestic cow], \elephant [African savanna elephant], \opossum [gray
short-tailed opossum], \galili [blind mole rat from Upper Galilee],
\carmeli [blind mole rat from northern Israel]). Information on
specimen ID, place and time of collection, and collector is shown for
each sample in \DLA{Supplemental Table 1}. In the case of \bonobo,
\cow, and \elephant, the placenta was collected at term upon expulsion
from the mother following delivery. Placental tissue from \human was
collected upon delivery via a caesarian section. Near-term placental
tissue from \dog was collected upon the death of the mother from a car
accident. Placental tissue from \opossum was sampled at fetal stage
32, roughly one day prior to birth. Only the fetal portion of the
placenta was sampled at the maternal-fetal interface (i.e. villous
portion for \human, \bonobo, \dog, and \elephant; cotyledon for \cow;
diffuse placenta for \opossum). All tissues were collected and stored
in RNAlater (Life Technologies, Carlsbad, CA) until RNA isolation
could be performed. Tissues were homogenized in Trizol, and total RNA
was isolated following the manufacturer's instructions. RNA was
further purified using RNeasy Midi Kit (Qiagen, Valencia, CA), in
conjunction with the RNase-Free DNase Set (Qiagen, Valencia, CA),
following the manufacturer’s protocol to remove any contaminating DNA
from the sample. Total RNA was then evaluated on a Nanodrop 1000
(Thermo Scientific, Wilmington, DE) and an Agilent Bioanalyzer 2100
(Santa Clara, CA) to confirm concentration and quality of the
isolation. RNA quality was assessed using A260/A280 and A260/A230
absorbance values.

\DLA{Add details for new sequences and fix numbers}

@@

\begin{figure}
<<species.tree,fig.width=8,fig.height=6,echo=FALSE>>=
species.tree <- 
    read.tree(text="(Monodelphis_domestica,((Dasypus_novemcinctus,Loxodonta_africana)Atlantogenata,((Canis_familiaris,(Equus_caballus,(Sus_scrofa,(Ovis_aries,Bos_taurus))))Laurasiatheria,(((Spalax_carmeli,Nannospalax_galili),Mus_Musculus)Rodentia,((Pan_paniscus,Homo_sapiens),Ateles_fuciceps)Primates)Euarchontoglires)Boreoeutheria)Eutheria);")
pp <- ggtree(species.tree)+
    geom_text(aes(label=label),hjust=0,size=5,fontface="italic")+
    scale_x_continuous(limits=c(0,10))
print(pp %>%
      ## monodelphis
      phylopic("91324e57-b3f1-42e0-abe3-43e5bc8aa4c6", color="#86B875",alpha=0.5,node=1,width=0.2)
      ## Armadillo
      %>% phylopic("5d59b5ce-c1dd-40f6-b295-8d2629b9775e", color="#44EEBB",alpha=0.5,node=2,width=0.18)
      ## elephant
      %>% phylopic("80db1004-bc9f-4318-84e7-cdd9639a1f3e",color="#BB1111",alpha=0.5,node=3,width=0.18)
      ## dog
      %>% phylopic("4d83a0cd-cf06-4a32-9a5a-0a6b644158c1",color="#FFAACC",alpha=0.5,node=4,width=0.19)
      ## horse
      %>% phylopic("071ee517-a0f1-4d19-aa29-812b9f86cb53",color="#22BBBB",alpha=0.5,node=5,width=0.18)
      ## pig
      %>% phylopic("008d6d88-d1be-470a-8c70-73625c3fb4fb",color="#33AA11",alpha=0.5,node=6,width=0.15)
      ## sheep
      %>% phylopic("cbea1964-89bf-4ce1-99b0-58fe0f9170b3",color="#004411",alpha=0.5,node=7,width=0.2)
      ## cow
      %>% phylopic("415714b4-859c-4d1c-9ce0-9e1081613df7",color="#881111",alpha=0.5,node=8,width=0.15)
      ## s carmeli
      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#881188",alpha=0.5,node=9,width=0.15)
      ## n galili
      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#888888",alpha=0.5,node=10,width=0.15)
      ## mouse
      %>% phylopic("6b2b98f6-f879-445f-9ac2-2c2563157025",color="#BB11CC",alpha=0.5,node=11,width=0.2)
      ## bonobo
      %>% phylopic("f06b4dd0-094e-4f37-9105-25fdb4eb1b02",color="#1111FF",alpha=0.5,node=14,width=0.18)
      ## spider monkey
      %>% phylopic("aceb287d-84cf-46f1-868c-4797c4ac54a8",color="#FF11FF",alpha=0.5,node=12,width=0.25)
      ## human
      %>% phylopic("a9f4ebd5-53d7-4de9-9e66-85ff6c2d513e",color="#CCCC44",alpha=0.5,node=13,width=0.18)
      )
@ 
\caption{Species tree for organisms studied}
\label{fig:species_tree}
\end{figure}

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of \bonobo are shown in \cref{tab:top50_bonobo_placenta}
with a list of all genes in Supplemental
\cref{tab:all_pan_paniscus_placenta}. Expression in the \bonobo
placenta differs somewhat from that of \human, as evidenced by the
makeup of the top 50 genes (\cref{tab:top50_bonobo_placenta}). The highest expression
 in \bonobo is %
\Sexpr{gene.expression("pan paniscus","PRG2",complete=TRUE)}, %
%
which is involved in insulin growth factor signaling and angiogenesis
during pregnancy \cite{Weyer.ea2011:Placentalregulationofpeptide}.
\textit{PRG2} is also expressed at relatively high levels (though
comparatively much lower) in the human placenta
(\Sexpr{gene.expression("homo sapiens","PRG2")};
Supplemental \cref{tab:all_homo_sapiens_placenta}). Four \bonobo genes
related to the chorionic somatomammotropins
are also expressed at relatively high levels (GH1 
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034560",complete=TRUE)}, 
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034212",complete=TRUE)}, 
GH2
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000030805",complete=TRUE)}, 
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034209",complete=TRUE)}, 
%
\cref{tab:all_pan_paniscus_placenta}, \cref{fig:csh1_evolution_tree}).
Placental expression of the genes involved in estrogen and
progesterone production (\textit{CYP19A1}, \textit{HSD3B1}
respectively) show lower expression in \bonobo as compared to \human %
(\Sexpr{gene.expression("pan paniscus","CYP19A1",complete=TRUE)},
\cref{fig:estrogen_evolution_tree}; %
\Sexpr{gene.expression("pan paniscus","HSD3B1",complete=TRUE)},
\cref{fig:progesterone_evolution_tree} and
\cref{tab:top50_human_placenta}) although the expression of
\textit{HSD3B1} is fairly similar to that of \smonkey %
(\Sexpr{gene.expression("ateles fusciceps","HSD3B1",complete=TRUE)}. %
In addition, \textit{CGA} is expressed in the bonobo placenta %
(\Sexpr{gene.expression("pan paniscus","CGA")}), but at much
lower levels than \human as well (\cref{tab:top50_human_placenta}).
%
Two Ensembl-annotated beta subunits of the chorionic gonadotropin
genes are expressed in \bonobo, one related to \textit{CGB} and
another related to \textit{CGB7}; both are not highly expressed
(\cref{fig:cgb_evolution_tree}). Among the other top 50 expressed
genes, there is high expression of keratins (e.g., \textit{KRT19},
\textit{KRT18}), heat-shock proteins (e.g., \textit{HSP90B1},
\textit{HSPA1A}), serpins (\textit{SERPINE1}, \textit{SERPINE2},
Supplemental \cref{fig:serpine_evolution_tree}), insulin growth factor
2 (\textit{IGF2}) and one of its binding proteins \textit{IGFBP3}, and
the hormone adrenomedullin (\textit{ADM}).

@@

The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
in the term placenta of \mouse are shown in
\cref{tab:top50_mouse_placenta} with a list of all genes with
$\mathrm{FPKM} \ge 50$ in Supplemental
\cref{tab:all_mus_musculus_placenta}. The top expressed genes include
\Sexpr{gene.expression("mus_musculus","Mir6236",complete=TRUE)}, whose
target \textit{in silico} is the orphan G-coupled receptor
\textit{Mrgprg}
\cite{Agarwal.ea2015:PredictingeffectivemicroRNAtarget},
\Sexpr{gene.expression("mus_musculus","H19",complete=TRUE)} and
\Sexpr{gene.expression("mus_musculus","Gm27786",complete=TRUE)}, both
of which are maternally imprinted (the later of which is a homolog of
human \textit{H19\_2}),
\Sexpr{gene.expression("mus_musculus","Igf2",complete=TRUE)}, numerous
cathepsins and a rodentia specific expansion of prolactin related
genes. There are also two miRNA which overlap with the of the 7S
signal recognition particle
(\Sexpr{gene.expression("mus_musculus","AC157822.1",complete=TRUE)}
and
\Sexpr{gene.expression("mus_musculus","AC099934.1",complete=TRUE)})
and numerous pseudogenes
(\Sexpr{gene.expression("mus_musculus","Gm11966",complete=TRUE)}
\Sexpr{gene.expression("mus_musculus","Gm10736",complete=TRUE)}
\Sexpr{gene.expression("mus_musculus","Gm10180",complete=TRUE)}
\Sexpr{gene.expression("mus_musculus","Gm29216",complete=TRUE)}) which
appear to be highly expressed, however, the short read length (34--44)
in many samples is likely preventing unambiguous assignment of all
reads to the more lengthy complete transcript, so these shorter
transcripts should be treated with suspicion, and we will ignore them
in further discussion. The prolactin-related genes
(\Sexpr{gene.expression("mus_musculus","Prl3b1",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Prl2b1",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Prl8a9",complete=TRUE)}, and
\Sexpr{gene.expression("mus_musculus","Prl8a6",complete=TRUE)}) are
very highly expressed in \mouse in, with %
\Sexpr{combined.fpkm[species=="mus musculus" &
  grepl("Prl(\\d.\\d)?",gene_short_name), sum(mean_fpkm >= 1)]}
prolactin related genes having $\mathrm{FPKM} \ge 1$
(\cref{fig:prolactin_evolution_tree}, Supplemental
\cref{fig:prolactin_receptor_evolution_tree}). The cathespins have
also expanded in rodents, with six showing high expression
(\Sexpr{gene.expression("mus_musculus","Ctsj",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Cts6",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Ctsr",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Ctsq",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Cts3",complete=TRUE)}, and
\Sexpr{gene.expression("mus_musculus","Ctsm",complete=TRUE)},
\cref{tab:top50_mouse_placenta,fig:cathespin_evolution_tree}).
Finally, \textit{Peg10}, an imprinted gene required for proper
formation of the placenta in mice is also highly expressed
(\Sexpr{gene.expression("mus_musculus","Peg10")}), as previously
reported \cite{Ono.ea2006:DeletionofPeg10imprinted}.

@@

\begin{table}[h]
  \footnotesize
<<top.50.horse.expressed.genes,echo=FALSE,results="asis">>=
top.horse.genes <- top.genes.table(species.name="equus caballus")
print(xtable(top.horse.genes),
      booktabs=TRUE,
      floating=FALSE,
      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
setkey(top.horse.genes,"Gene symbol")
@ 
\caption{\tableexplanation{\horse}{tab:all_equus_caballus_placenta}}
\label{tab:top50_horse_placenta}
\end{table}

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of \elephant are shown in \cref{tab:top50_elephant_placenta}
with a list of all genes in
\cref{tab:all_loxodonta_africana_placenta}. The gene with the highest
expression was the natriuretic peptide precursor B
(\Sexpr{gene.expression("loxodonta_africana","NPPB",complete=TRUE)}).
Two thymosins,
\Sexpr{gene.expression("loxodonta_africana","TMSB10",complete=TRUE)}
and
\Sexpr{gene.expression("loxodonta_africana","TMSB4Y",complete=TRUE)}
were also also was highly expressed. The miRNA mir-675
(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000027406",complete=TRUE)}).
which is imbedded in the first exon of the maternally imprinted long
non-coding RNA gene \textit{H19}
\cite{Keniry.ea2012:H19lincRNAisdevelopmental}, a gene which is
currently un-annotated in \elephant{}. \DLA{Probably consider fixing
  the unannotation of H19; it seems pretty obvious to me, though it
  would require fixing the GTF file for elephant with the right
  \href{http://useast.ensembl.org/Loxodonta_africana/Transcript/Summary?db=core;pt=GENSCAN00000034833;r=scaffold_145:855987-899506}{genscan
    prediction} for this region} Five uncharacterized genes are
expressed to a high degree in the \elephant placenta
(\cref{tab:top50_elephant_placenta}), three of which
(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000028493",complete=TRUE)},
\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000032011",complete=TRUE)},
\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000031933",complete=TRUE)})
are members of an \elephant{}-specific expansion of 11 genes closely
related to the thymus-specific serine protease gene \textit{PRSS16}
\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~C The other 8 \textit{PRSS16}-like
genes are also expressed to a lesser degree
(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000026505",complete=TRUE)}).
Two other novel genes with high expression are related to the gene
\textit{SSLP1}
(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000032659",complete=TRUE)}
and
\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000029551",complete=TRUE)},
\cref{fig:sslp1_evolution_tree}), which is also expressed in \cow and
\dog placentas (\cref{tab:top50_dog_placenta,tab:top50_cow_placenta}).
Placental expression of the hormones relaxin (\textit{RLN}, %
\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000036090")})
and prolactin (\textit{PRL}) are exceptionally high in the elephant
when compared to other mammals
(\cref{fig:rln2_evolution_tree,fig:prolactin_evolution_tree}). There
is little to no placental expression of the estrogen synthase gene,
\textit{CYP19A1}
(\Sexpr{gene.expression("loxodonta_africana","CYP19A1")}) and the
enzyme that produces progesterone HSD3B1
(\Sexpr{gene.expression("loxodonta_africana","HSD3B")}). Other genes
with high expression in the elephant placenta include genes involved
in oxidative stress
(\Sexpr{gene.expression("loxodonta_africana","TXN",complete=TRUE)}),
growth
(\Sexpr{gene.expression("loxodonta_africana","IGF2",complete=TRUE)})
and
\Sexpr{gene.expression("loxodonta_africana","PAEP",complete=TRUE)}, a
gene usually expressed in the endometrium of the uterus in humans
\cite{Rutanen.ea1987:Progesterone-associatedproteinsPP12and}.

@@

\begin{table}
  \footnotesize
<<go.analysis.all.tissue,echo=FALSE,results="asis">>=
go.analysis.wide <- 
    data.table(dcast(data=go.results,
                     formula=GO.ID+type+Term+Ontology~species,
                     value.var="FDR"))
go.analysis.wide <- 
    go.analysis.wide[,`n(FDR ≤ 0.05)`:=apply(go.analysis.wide[,-(1:4),with=FALSE],1,
                                         function(x){sum(as.numeric(as.character(x)) <= 0.05)})]
go.analysis.wide <- 
    go.analysis.wide[,`max(FDR ≤ 0.05)`:=apply(go.analysis.wide[,-(1:4),with=FALSE],1,
                                           function(x){if (!any(is.finite(x) & x<=0.05)) {
                                                           return(NA)
                                                       } else {
                                                           max(x[is.finite(x) & x<=0.05],na.rm=TRUE)
                                                       }})]
go.analysis.wide.table <- 
    go.analysis.wide[!is.na(`n(FDR ≤ 0.05)`) & `n(FDR ≤ 0.05)` >= 12 &
                         grepl("Top 1% No Housekeeping",type),
                     list(GO.ID,Term,Ontology,`n(FDR ≤ 0.05)`,`max(FDR ≤ 0.05)`)]
go.analysis.wide.table <- 
    go.analysis.wide.table[order(Ontology,-`n(FDR ≤ 0.05)`,`max(FDR ≤ 0.05)`)]
setnames(go.analysis.wide.table,"GO.ID","GO ID")

print(xtable(go.analysis.wide.table,
             display=c("d","s","s","s","d","e"),
             align=  c("c","c","c","c","c","c")
             ),
      math.style.exponents=TRUE,
      include.rownames=FALSE,
      booktabs=TRUE,
      floating=FALSE
      )

@ 
\caption{GO terms with significant enrichment ($\mathrm{FDR}\le0.05$)
  in 12 or more of the species analyzed from genes in the top 1\% of
  expression. $n(\mathrm{FDR}\le0.05)$ is the number of species with
  significant enrichment ($\mathrm{FDR}\le 0.05$) in the GO term
  listed. $max(\mathrm{FDR}\le0.05)$ is the maximum FDR value in all
  species with an FDR of less than $0.05$.}
\label{tab:go_terms_all_species}
\end{table}

@@

\subsubsection{\Ac{GH} family}

>>>>

\newcommand{\genetreeexplanation}[1]{Gene tree and heatmap of
  expression of genes related to #1. The gene tree was obtained from
  Ensembl~\cite{Cunningham.ea2015:Ensembl2015} and genes from species
  not included in this study or having expression by \ac{FPKM} less
  than 1 were excluded from the plotted tree. Heatmap cells in white
  are from species where there was no expression of the relevant
  transcript (or a species where that transcript does not apply).}

@@

<<plot.defaults,echo=FALSE,results="hide">>=
grad.def <- list(low = "#132B43", high = "#56B1F7", space = "Lab", na.value = NA)

@@

Fetal tissue was collected from the placenta of eight mammalian
species (\human [human], \bonobo [bonobo], \dog [domestic dog], \cow
[domestic cow], \elephant [African savanna elephant], \opossum [gray
short-tailed opossum], \galili [blind mole rat from Upper Galilee],
\carmeli [blind mole rat from northern Israel]). Information on
specimen ID, place and time of collection, and collector is shown for
each sample in Supplemental \cref{tab:srr_numbers}. In the case of \bonobo,
\cow, and \elephant, the placenta was collected at term upon expulsion
from the mother following delivery. Placental tissue from \human was
collected upon delivery via a caesarian section. Near-term placental
tissue from \dog was collected upon the death of the mother from a car
accident. Placental tissue from \opossum was sampled at fetal stage
32, roughly one day prior to birth. Only the fetal portion of the
placenta was sampled at the maternal-fetal interface (i.e. villous
portion for \human, \bonobo, \dog, and \elephant; cotyledon for \cow;
diffuse placenta for \opossum). All tissues were collected and stored
in RNAlater (Life Technologies, Carlsbad, CA) until RNA isolation
could be performed. Tissues were homogenized in Trizol, and total RNA
was isolated following the manufacturer's instructions. RNA was
further purified using RNeasy Midi Kit (Qiagen, Valencia, CA), in
conjunction with the RNase-Free DNase Set (Qiagen, Valencia, CA),
following the manufacturer’s protocol to remove any contaminating DNA
from the sample. Total RNA was then evaluated on a Nanodrop 1000
(Thermo Scientific, Wilmington, DE) and an Agilent Bioanalyzer 2100
(Santa Clara, CA) to confirm concentration and quality of the
isolation. RNA quality was assessed using A260/A280 and A260/A230
absorbance values.

% \DLA{Add details for new sequences and fix numbers}

@@

\begin{figure}
<<species.tree,fig.width=8,fig.height=6,echo=FALSE>>=
species.tree <- 
    read.tree(text="(Monodelphis_domestica,((Dasypus_novemcinctus,Loxodonta_africana)Atlantogenata,((Canis_familiaris,(Equus_caballus,(Sus_scrofa,(Ovis_aries,Bos_taurus))))Laurasiatheria,(((Spalax_carmeli,Nannospalax_galili),Mus_Musculus)Rodentia,((Pan_paniscus,Homo_sapiens),Ateles_fuciceps)Primates)Euarchontoglires)Boreoeutheria)Eutheria);")
pp <- ggtree(species.tree)+
    geom_text(aes(label=label),hjust=0,size=5,fontface="italic")+
    scale_x_continuous(limits=c(0,10))
print(pp %>%
      ## monodelphis
      phylopic("91324e57-b3f1-42e0-abe3-43e5bc8aa4c6", color="#86B875",alpha=0.5,node=1,width=0.2)
      ## Armadillo
      %>% phylopic("5d59b5ce-c1dd-40f6-b295-8d2629b9775e", color="#44EEBB",alpha=0.5,node=2,width=0.15)
      ## elephant
      %>% phylopic("80db1004-bc9f-4318-84e7-cdd9639a1f3e",color="#BB1111",alpha=0.5,node=3,width=0.15)
      ## dog
      %>% phylopic("4d83a0cd-cf06-4a32-9a5a-0a6b644158c1",color="#FFAACC",alpha=0.5,node=4,width=0.15)
      ## horse
      %>% phylopic("071ee517-a0f1-4d19-aa29-812b9f86cb53",color="#22BBBB",alpha=0.5,node=5,width=0.15)
      ## pig
      %>% phylopic("008d6d88-d1be-470a-8c70-73625c3fb4fb",color="#33AA11",alpha=0.5,node=6,width=0.12)
      ## sheep
      %>% phylopic("cbea1964-89bf-4ce1-99b0-58fe0f9170b3",color="#004411",alpha=0.5,node=7,width=0.18)
      ## cow
      %>% phylopic("415714b4-859c-4d1c-9ce0-9e1081613df7",color="#881111",alpha=0.5,node=8,width=0.12)
      ## s carmeli
      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#881188",alpha=0.5,node=9,width=0.1)
      ## n galili
      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#888888",alpha=0.5,node=10,width=0.1)
      ## mouse
      %>% phylopic("6b2b98f6-f879-445f-9ac2-2c2563157025",color="#BB11CC",alpha=0.5,node=11,width=0.18)
      ## bonobo
      %>% phylopic("f06b4dd0-094e-4f37-9105-25fdb4eb1b02",color="#1111FF",alpha=0.5,node=12,width=0.18)
      ## spider monkey
      %>% phylopic("aceb287d-84cf-46f1-868c-4797c4ac54a8",color="#FF11FF",alpha=0.5,node=14,width=0.25)
      ## human
      %>% phylopic("a9f4ebd5-53d7-4de9-9e66-85ff6c2d513e",color="#CCCC44",alpha=0.5,node=13,width=0.18)
      )
@ 
\caption{Species tree for organisms studied}
\label{fig:species_tree}
\end{figure}

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of \bonobo are shown in \cref{tab:top50_bonobo_placenta}
with a list of all genes in Supplemental
\cref{tab:all_pan_paniscus_placenta}. Expression in the \bonobo
placenta differs somewhat from that of \human, as evidenced by the
makeup of the top 50 genes (\cref{tab:top50_bonobo_placenta}). The highest expression
 in \bonobo is %
\Sexpr{gene.expression("pan paniscus","PRG2",complete=TRUE)}, %
%
which is involved in insulin growth factor signaling and angiogenesis
during pregnancy \cite{Weyer.ea2011:Placentalregulationofpeptide}.
\textit{PRG2} is also expressed at relatively high levels (though
comparatively much lower) in the human placenta
(\Sexpr{gene.expression("homo sapiens","PRG2")};
Supplemental \cref{tab:all_homo_sapiens_placenta}). Four \bonobo genes
related to the chorionic somatomammotropins
are also expressed at relatively high levels (GH1 
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034560",complete=TRUE)}, 
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034212",complete=TRUE)}, 
GH2
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000030805",complete=TRUE)}, 
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034209",complete=TRUE)}, 
%
\cref{tab:all_pan_paniscus_placenta}, \cref{fig:csh1_evolution_tree}).
Placental expression of the genes involved in estrogen and
progesterone production (\textit{CYP19A1}, \textit{HSD3B1}
respectively) show lower expression in \bonobo as compared to \human %
(\Sexpr{gene.expression("pan paniscus","CYP19A1",complete=TRUE)},
\cref{fig:estrogen_evolution_tree}; %
\Sexpr{gene.expression("pan paniscus","HSD3B1",complete=TRUE)},
\cref{fig:progesterone_evolution_tree} and
\cref{tab:top50_human_placenta}) although the expression of
\textit{HSD3B1} is fairly similar to that of \smonkey %
(\Sexpr{gene.expression("ateles fusciceps","HSD3B1",complete=TRUE)}. %
In addition, \textit{CGA} is expressed in the bonobo placenta %
(\Sexpr{gene.expression("pan paniscus","CGA")}), but at much
lower levels than \human as well (\cref{tab:top50_human_placenta}).
%
Two Ensembl-annotated beta subunits of the chorionic gonadotropin
genes are expressed in \bonobo, one related to \textit{CGB} and
another related to \textit{CGB7}; both are not highly expressed
(\cref{fig:cga_cgb_evolution_tree}). Among the other top 50 expressed
genes, there is high expression of keratins (e.g., \textit{KRT19},
\textit{KRT18}), heat-shock proteins (e.g., \textit{HSP90B1},
\textit{HSPA1A}), serpins (\textit{SERPINE1}, \textit{SERPINE2},
Supplemental \cref{fig:serpine_evolution_tree}), insulin growth factor
2 (\textit{IGF2}) and one of its binding proteins \textit{IGFBP3}, and
the hormone adrenomedullin (\textit{ADM}).

@@

The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
in the term placenta of \mouse are shown in
\cref{tab:top50_mouse_placenta} with a list of all genes with
$\mathrm{FPKM} \ge 50$ in Supplemental
\cref{tab:all_mus_musculus_placenta}. The top expressed genes include
\Sexpr{gene.expression("mus_musculus","Mir6236",complete=TRUE)}, whose
target \textit{in silico} is the orphan G-coupled receptor
\textit{Mrgprg}
\cite{Agarwal.ea2015:PredictingeffectivemicroRNAtarget},
\Sexpr{gene.expression("mus_musculus","H19",complete=TRUE)} and
\Sexpr{gene.expression("mus_musculus","Gm27786",complete=TRUE)}, both
of which are maternally imprinted (the later of which is a homolog of
human \textit{H19\_2}),
\Sexpr{gene.expression("mus_musculus","Igf2",complete=TRUE)}, numerous
cathepsins and a rodentia specific expansion of prolactin related
genes. There are also two miRNA which overlap with the of the 7S
signal recognition particle
(\Sexpr{gene.expression("mus_musculus","AC157822.1",complete=TRUE)}
and
\Sexpr{gene.expression("mus_musculus","AC099934.1",complete=TRUE)})
and numerous pseudogenes
(\Sexpr{gene.expression("mus_musculus","Gm11966",complete=TRUE)}
\Sexpr{gene.expression("mus_musculus","Gm10736",complete=TRUE)}
\Sexpr{gene.expression("mus_musculus","Gm10180",complete=TRUE)}
\Sexpr{gene.expression("mus_musculus","Gm29216",complete=TRUE)}) which
appear to be highly expressed, however, the short read length (34--44)
in many samples is likely preventing unambiguous assignment of all
reads to the more lengthy complete transcript, so these shorter
transcripts should be treated with suspicion, and we will ignore them
in further discussion. The prolactin-related genes
(\Sexpr{gene.expression("mus_musculus","Prl3b1",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Prl2b1",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Prl8a9",complete=TRUE)}, and
\Sexpr{gene.expression("mus_musculus","Prl8a6",complete=TRUE)}) are
very highly expressed in \mouse in, with %
\Sexpr{combined.fpkm[species=="mus musculus" &
  grepl("Prl(\\d.\\d)?",gene_short_name), sum(mean_fpkm >= 1)]}
prolactin related genes having $\mathrm{FPKM} \ge 1$
(\cref{fig:prolactin_evolution_tree}, Supplemental
\cref{fig:prolactin_receptor_evolution_tree}). The cathespins have
also expanded in rodents, with six showing high expression
(\Sexpr{gene.expression("mus_musculus","Ctsj",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Cts6",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Ctsr",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Ctsq",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Cts3",complete=TRUE)}, and
\Sexpr{gene.expression("mus_musculus","Ctsm",complete=TRUE)},
\cref{tab:top50_mouse_placenta,fig:pag2_prss16_ctsl_psg_evolution_tree}).
Finally, \textit{Peg10}, an imprinted gene required for proper
formation of the placenta in mice is also highly expressed
(\Sexpr{gene.expression("mus_musculus","Peg10")}), as previously
reported \cite{Ono.ea2006:DeletionofPeg10imprinted}.

@@

\begin{table}[h]
  \footnotesize
<<top.50.horse.expressed.genes,echo=FALSE,results="asis">>=
output.top.genes.table(top.genes.table(species.name="equus caballus"))
@ 
\caption{\tableexplanation{\horse}{tab:all_equus_caballus_placenta}}
\label{tab:top50_horse_placenta}
\end{table}

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of \elephant are shown in \cref{tab:top50_elephant_placenta}
with a list of all genes in
\cref{tab:all_loxodonta_africana_placenta}. The gene with the highest
expression was the natriuretic peptide precursor B
(\Sexpr{gene.expression("loxodonta_africana","NPPB",complete=TRUE)}).
Two thymosins,
\Sexpr{gene.expression("loxodonta_africana","TMSB10",complete=TRUE)}
and
\Sexpr{gene.expression("loxodonta_africana","TMSB4Y",complete=TRUE)}
were also also was highly expressed. The miRNA mir-675
(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000027406",complete=TRUE)}).
which is imbedded in the first exon of the maternally imprinted long
non-coding RNA gene \textit{H19}
\cite{Keniry.ea2012:H19lincRNAisdevelopmental}, a gene which is
currently un-annotated in \elephant{}. %
% \DLA{Probably consider fixing the unannotation of H19; it seems pretty
%   obvious to me, though it would require fixing the GTF file for
%   elephant with the right
%   \href{http://useast.ensembl.org/Loxodonta_africana/Transcript/Summary?db=core;pt=GENSCAN00000034833;r=scaffold_145:855987-899506}{genscan
%     prediction} for this region} %
Five uncharacterized genes are expressed to a high degree in the
\elephant placenta (\cref{tab:top50_elephant_placenta}), three of
which
(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000028493",complete=TRUE)},
\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000032011",complete=TRUE)},
\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000031933",complete=TRUE)})
are members of an \elephant{}-specific expansion of 11 genes closely
related to the thymus-specific serine protease gene \textit{PRSS16}
\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~C The other 8
\textit{PRSS16}-like genes are also expressed to a lesser degree
(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000026505",complete=TRUE)}).
Two other novel genes with high expression are related to the gene
\textit{SSLP1}
(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000032659",complete=TRUE)}
and
\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000029551",complete=TRUE)},
\cref{fig:sslp1_evolution_tree}), which is also expressed in \cow and
\dog placentas (\cref{tab:top50_dog_placenta,tab:top50_cow_placenta}).
Placental expression of the hormones relaxin (\textit{RLN}, %
\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000036090")})
and prolactin (\textit{PRL}) are exceptionally high in the elephant
when compared to other mammals
(\cref{fig:rln2_evolution_tree,fig:prolactin_evolution_tree}). There
is little to no placental expression of the estrogen synthase gene,
\textit{CYP19A1}
(\Sexpr{gene.expression("loxodonta_africana","CYP19A1")}) and the
enzyme that produces progesterone HSD3B1
(\Sexpr{gene.expression("loxodonta_africana","HSD3B")}). Other genes
with high expression in the elephant placenta include genes involved
in oxidative stress
(\Sexpr{gene.expression("loxodonta_africana","TXN",complete=TRUE)}),
growth
(\Sexpr{gene.expression("loxodonta_africana","IGF2",complete=TRUE)})
and
\Sexpr{gene.expression("loxodonta_africana","PAEP",complete=TRUE)}, a
gene usually expressed in the endometrium of the uterus in humans
\cite{Rutanen.ea1987:Progesterone-associatedproteinsPP12and}.

@@

@@

\subsubsection{\ac{GH} family}
